---
title: "Creating Sashimi-Style Plots with spliceTerrain"
author:
- name: Lachlan Baer
  affiliation:
  - Robinson Research Institute, Adelaide University, Adelaide, SA, Australia
  - College of Health, Adelaide University, Adelaide, SA, Australia
  email: bioc@lwwb.dev
package: spliceTerrain
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    css: custom.css
# bibliography: references.bib
vignette: >
    %\VignetteIndexEntry{Creating Sashimi-Style Plots with spliceTerrain}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r global-opts, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", message = FALSE, crop = NULL
)
if (!interactive())
    # Wide chunk output will horizontal scroll instead of wrap
    options(width = 999)
```

# Setup

`BiocManager` is recommended for the installation of packages required in this
vignette.
`BiocManager` handles the installation of packages from Bioconductor, CRAN and
GitHub repositories.

```{r install, eval=FALSE}
if (!"BiocManager" %in% rownames(installed.packages()))
    install.packages("BiocManager")
pkgs <- c(
    "spliceTerrain", "RNAseqData.HNRNPC.bam.chr14", "EnsDb.Hsapiens.v75",
    "ensembldb", "stringr", "tibble", "pander"
)
BiocManager::install(pkgs, update = FALSE)
```

Now we can load the required packages.

```{r pkgs}
library(spliceTerrain)
library(RNAseqData.HNRNPC.bam.chr14)
library(EnsDb.Hsapiens.v75)
library(ensembldb)
library(stringr)
library(tibble)
library(pander)
```

# Quick start

At a minimum, we require BAM file paths and a genomic interval to plot.

Here we use BAM files from the `RNAseqData.HNRNPC.bam.chr14` package as example data, selecting two of these files to showcase `spliceTerrain`.
In the most simple case, a genomic interval can be specified as a character vector of length 1.
That's all we need to build a simple sashimi plot!

```{r quickstart, fig.wide=TRUE, fig.height=6}
bams <- RNAseqData.HNRNPC.bam.chr14_BAMFILES[c(7,1)]
region <- "chr14:70,233,810-70,238,690"
spliceTerrain(bam = bams, region = region)
```

# Introduction

Sashimi plots are a visualisation tool for RNA-seq data that display summarised read coverage across a genomic interval and how splice junctions bridge exons.
By displaying both per-base coverage and junction-spanning reads, sashimi plots provide an intuitive view of transcript structure and enable assessment of exon usage, alternative splicing events, and isoform differences between samples.

## Motivation

`spliceTerrain` was designed to integrate seamlessly with the Bioconductor ecosystem, utilising core Bioconductor classes and methods at the centre of its workflow.
During RNA-seq analysis, genomic intervals, annotations, and features of interest are often represented as `GRanges` objects.
`spliceTerrain` leverages this to enable straightforward visualisation that integrates seamlessly into `Rmarkdown`/`Quarto` documents, while promoting reproducibility through a user-friendly R interface for generating sashimi plots within script-based analyses.
`spliceTerrain` enables publication-quality figures while retaining flexibility.
Users can optionally access the underlying plot data and individual plot panels, allowing further customisation or arrangement of panels.

## Previous works

Several established and powerful tools also exist for plotting sashimi-style visualisations, such as `MISO`, `ggsashimi` and `IGV`.
`spliceTerrain` complements these by offering the choice of a Bioconductor-native, highly configurable approach tailored to scientists that adopt an `R`-based workflow.

# Complete guide

## Required arguments

First, we load the minimal required objects needed to create a sashimi plot with `spliceTerrain`. The BAM file paths were defined above in the [Quick start](#quick-start) section, but are defined again below.

```{r bams}
bams <- RNAseqData.HNRNPC.bam.chr14_BAMFILES[c(7,1)]
```

Because these files were aligned to GRCh37, we will use an Ensembl annotation database based on the same genome build.

```{r ensDb}
ensDb <- EnsDb.Hsapiens.v75
```

We then use a transcript ID (ENST00000394366) to retrieve a single-range `GRanges` object, which will define the plotting region.
The `region` argument can also accept a `GRanges` object with multiple ranges, or even a `GRangesList.`
However, `spliceTerrain` will internally reduce this input to a single range spanning all supplied intervals.

The selected BAM files use UCSC-style sequence names, so the annotation must also be converted to the same naming convention before being supplied to `spliceTerrain`.

```{r region}
region <- transcripts(ensDb, filter = TxIdFilter("ENST00000394366"))
seqlevelsStyle(region) <- "ucsc"
```

## Adding transcript models

Transcript models can be displayed by supplying the `annotation` argument as a `GRangesList`, where each list element represents a single transcript.
Transcript labels plotted on the y-axis are taken from the names of the `GRangesList` elements, otherwise a placeholder will be used.
For this vignette, the plotting region covers the *SRSF5* gene, so we use its Ensembl gene ID to retrieve exon ranges for each transcript.
To keep the example compact and readable, we subset the result to the first seven transcripts.

```{r annotation}
annotation <- exonsBy(
    ensDb, by = "tx", filter = GeneIdFilter("ENSG00000100650")
)
annotation <- annotation[1:7]
seqlevelsStyle(annotation) <- "ucsc"
```

```{r annotation_plot, fig.wide=TRUE, fig.height=9}
spliceTerrain(bam = bams, region = region, annotation = annotation)
```

When an annotation track is included, the plot contains one additional panel.
In other words, the total number of panels becomes the number of BAM files plus one annotation panel.
This becomes important when using the `panel_heights` argument described below.

## Panel heights

The relative height of each plot panel can be controlled with the `panel_heights` argument.
When an annotation track is present, its height is always specified by the final value in the vector.
For example, to make the annotation track four times taller than each sample sashimi panel we can write:

```{r panel_heights, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation,
    panel_heights = c(1, 1, 4)
)
```

## Annotation labels

Labels can be added to exons in the annotation track by referring to a metadata column in the annotation object.
For example, this object already contains exon rank information for each exon.

```{r}
annotation[1]
```

To display these values on the annotation track, simply provide the name of the desired metadata column.

```{r anno_text_col, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation,
    anno_text_col = "exon_rank"
)
```

## Highlighting regions

To highlight a genomic region, we specify the interval to the `highlight` argument.
The `highlight_colour` argument can be used to specify the colour of the highlight.
Similar to the `region` argument, `highlight` accepts a `character` or `GRanges` object.
This can be used to specify a single base pair site, that may, for example, be used to highlight a splice site mutation.

```{r highlight, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation,
    highlight = "chr14:70234854-70234854", highlight_colour = "red"
)
```

Alternatively, we can highlight a wider range, which may be used to draw our attention to an alternatively used exon.

```{r highlight2, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation,
    highlight = "chr14:70234056-70234097",
    highlight_colour = scales::alpha("red", 0.2)
)
```

## Calculating PSI

Sashimi plots can be informative for visualising the usage of alternative splicing options.
Sometimes it is 

```{r psi, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation,
    psi = "chr14:70234854-70234854"
)
```

## Customizing arcs

```{r arcs, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation,
    psi = "chr14:70234854-70234854", arc_height = 0.4, scale_arcs = TRUE
)
```

## Colours

```{r colours, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation,
    colours = c("darkblue", "darkred")
)
```

## Compressing introns

```{r compress_introns, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation,
    compress_introns = TRUE, intron_width = 500
)
```

## Strandedness

```{r strandedness, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation,
    strandedness = "reverse"
)
```

## Minimum coverage

```{r min_coverage, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation,
    strandedness = "reverse", min_coverage = 5, min_junction_reads = 5
)
```

## Returning plot data

This way you can arrange the plots as you like for publications, investigate junction level data to find the psi that you want to calculate PSI values for, troubleshoot, etc.

# Advanced use cases

- Isolating a specific splice event
    * return_ctx, subsetByOverlaps, provide ctx arg
- Rearrangement of plots with patchwork, accessing individual plots

# Session Information

```{r session_info}
sessionInfo() |>
    pander()
```
