---
title: "Creating Sashimi-Style Plots with spliceTerrain"
author:
- name: Lachlan Baer
  affiliation:
  - Robinson Research Institute, Adelaide University, Adelaide, SA, Australia
  - College of Health, Adelaide University, Adelaide, SA, Australia
  email: bioc@lwwb.dev
package: spliceTerrain
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    css: custom.css
# bibliography: references.bib
vignette: >
    %\VignetteIndexEntry{Creating Sashimi-Style Plots with spliceTerrain}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r global-opts, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", message = FALSE, crop = NULL
)
if (!interactive())
    # Wide chunk output will horizontal scroll instead of wrap
    options(width = 999)
```

# Setup

`BiocManager` is recommended for installing the packages required in this
vignette.
`BiocManager` handles installation from Bioconductor, CRAN and GitHub
repositories.

```{r install, eval=FALSE}
if (!"BiocManager" %in% rownames(installed.packages()))
    install.packages("BiocManager")
pkgs <- c(
    "spliceTerrain", "RNAseqData.HNRNPC.bam.chr14", "EnsDb.Hsapiens.v75",
    "AnnotationFilter", "ensembldb", "GenomicRanges", "scales", "pander",
    "GenomeInfoDb"
)
BiocManager::install(pkgs, update = FALSE)
```

Now we can load the packages used below.

```{r pkgs}
library(spliceTerrain)
library(RNAseqData.HNRNPC.bam.chr14)
library(EnsDb.Hsapiens.v75)
library(AnnotationFilter)
library(ensembldb)
library(GenomicRanges)
library(scales)
library(pander)
library(GenomeInfoDb)
```

# Quick start

At a minimum, `spliceTerrain()` needs BAM file paths and a genomic interval.
Here we use BAM files from `RNAseqData.HNRNPC.bam.chr14`, selecting one HNRNPC
knockdown sample and one control sample.
The region can be supplied as a single character string.

```{r quickstart, fig.wide=TRUE, fig.height=6}
bams <- RNAseqData.HNRNPC.bam.chr14_BAMFILES[c(7, 1)]
region <- "chr14:70,233,810-70,238,690"

spliceTerrain(bam = bams, region = region)
```

# Introduction

Sashimi plots visualise RNA-seq alignments across a genomic interval.
They combine per-base read coverage with arcs representing reads that span
splice junctions.
This makes them useful for inspecting transcript structure, exon usage,
alternative splice sites, intron retention, and differences in junction usage
between samples.

## Motivation

Several excellent tools already exist for generating sashimi-style
visualisations, including genome browsers, command-line tools, and
workflow-specific plotting utilities.
`spliceTerrain` is intended to complement these approaches by providing an
R-native interface that works naturally within the Bioconductor ecosystem and
reproducible analysis documents.

In many RNA-seq analyses, regions of interest, annotations, and genomic features
are already represented as `GRanges` or `GRangesList` objects.
`spliceTerrain` accepts these objects directly, making it straightforward to
move from a differential splicing result or genomic annotation to a sashimi plot
inside the same R session.
For users who want more control, processed coverage and junction data can also
be returned, inspected, modified, and replotted.

# Complete guide

The following sections build up aspects of a more complete plot and introduce
the main options available in `spliceTerrain()`.
The examples continue to use the HNRNPC chr14 BAM files from the quick start.

## Input BAM files and regions

The `bam` argument is a character vector of indexed BAM file paths.
If the vector is named, those names are used as sample labels in the plot.
If it is unnamed, labels are derived from the BAM file names.
BAM files must be indexed for region-restricted import.
Index files should be discoverable by `Rsamtools` using standard same-directory
naming, typically `file.bam.bai`.

```{r named_bams}
bams <- RNAseqData.HNRNPC.bam.chr14_BAMFILES[c(7, 1)]
names(bams) <- c("HNRNPC knockdown", "Control")
```

The `region` argument can be a character string, a `GRanges`, or a
`GRangesList`.
Character regions may use comma-separated coordinates, such as
`"chr14:70,233,810-70,238,690"`.
For Bioconductor workflows, using a `GRanges` object is often more convenient.

Because the example BAM files use UCSC style chromosome names, we also convert
the region to UCSC style before plotting.

```{r region}
ensDb <- EnsDb.Hsapiens.v75
region <- transcripts(ensDb, filter = TxIdFilter("ENST00000394366"))
seqlevelsStyle(region) <- "UCSC"
region
```

If multiple ranges are supplied, `spliceTerrain()` reduces them to a single span
on one chromosome and uses that span as the BAM query and plotting window.
The span is the smallest continuous interval containing all supplied ranges: it
starts at the lowest start coordinate and ends at the highest end coordinate.

## Adding transcript annotation

Transcript models can be displayed with the `annotation` argument.
This must be a `GRangesList`, where each list element represents one transcript
or feature group.
The names of the list elements are used as y-axis labels in the annotation
panel.
When annotation is present, the final plot contains one panel per BAM file plus
one annotation panel.
The relative heights of these panels are controlled with `panel_heights`, where
the final value corresponds to the annotation panel.
Setting `annotated_junctions = TRUE` distinguishes junctions that exactly match
an intron in the supplied annotation with solid arcs, while unmatched junctions
are drawn with dashed arcs.
Junction classification is relative to the supplied annotation, so known
junctions may appear unmatched when only a subset of transcripts is provided.

For this vignette, the plotting region covers the *SRSF5* gene.
We retrieve exon ranges for the gene and keep a small number of transcripts so
the annotation panel remains readable.

```{r annotation}
annotation <- exonsBy(
    ensDb, by = "tx", filter = GeneIdFilter("ENSG00000100650")
)
annotation <- annotation[1:3]
seqlevelsStyle(annotation) <- "UCSC"
```

```{r annotation_plot, fig.wide=TRUE, fig.height=7.5}
spliceTerrain(
    bam = bams,
    region = region,
    annotation = annotation,
    annotated_junctions = TRUE,
    panel_heights = c(2, 2, 1)
)
```

## Styling annotation tracks

Annotation features can be coloured by a metadata column with `anno_fill_by`.
Use `"group"` to colour each transcript or feature group separately.
Specific fill colours can be supplied with `anno_fill_colours`.

Text labels can be drawn inside annotation features with `anno_label_by`.
The label colour and size are controlled by `anno_label_colour` and
`anno_label_size`.
The `min_arrow` argument controls how wide an annotation intron must be before
directional arrows are drawn.

```{r annotation_styling, fig.wide=TRUE, fig.height=7.5}
spliceTerrain(
    bam = bams,
    region = region,
    annotation = annotation,
    panel_heights = c(2, 2, 1),
    anno_fill_by = "group",
    anno_fill_colours = c(
        ENST00000553369 = "#66C2A5",
        ENST00000394366 = "#8DA0CB",
        ENST00000451983 = "#FC8D62"
    ),
    anno_label_by = "exon_rank",
    anno_label_colour = "black",
    anno_label_size = 2.5,
    min_arrow = 100
)
```

## Highlighting regions of interest

The `highlight` argument marks one or more genomic intervals with shaded
vertical bands.
This is useful for drawing attention to a splice site, variant position,
alternative exon, or other feature of interest.
The highlighted interval can be supplied as a character string or as a
`GRanges` object.

```{r highlight_site, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    highlight = "chr14:70234854-70234854",
    highlight_colour = "red"
)
```

A wider interval can be highlighted with a semi-transparent colour.

```{r highlight_exon, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    highlight = "chr14:70234056-70234097",
    highlight_colour = scales::alpha("red", 0.2)
)
```

## Percent spliced in labels

Percent spliced in (PSI) is commonly used to summarise how often a splice choice
is used relative to other local splice choices.
In `spliceTerrain()`, the `psi` argument identifies a local genomic interval and
labels junctions whose start or end anchor overlaps that interval.
Junction anchors use intronic coordinates: the start anchor is the first base of
the intron, and the end anchor is the last base of the intron, not the adjacent
exonic bases.
The label reports each selected junction's fraction of the included junction
reads.

This is intended as an interpretable local junction-usage label on the sashimi
plot, rather than full transcript-level isoform inference.
By default, the junction count and PSI percentage are separated by a new line.
Use `psi_label_sep = " "` to display them together on one line.

```{r psi, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    psi = "chr14:70234854-70234854",
    psi_label_sep = " "
)
```

## Filtering signal

Three arguments control how much signal is imported and displayed.
`min_mapq` filters alignments by mapping quality during BAM import.
`min_coverage` filters per-base coverage positions before plotting.
`min_junction_reads` filters splice junction arcs by their supporting read
count.
The latter two arguments also accept percentage strings such as `"10%"`.
Percentage thresholds are calculated separately for each BAM relative to its
maximum raw coverage or junction count within the plotting region.
When supplying one threshold per BAM, whole-number counts and percentages can be
mixed in a character vector.

```{r filtering, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    min_mapq = 10,
    min_coverage = 10,
    min_junction_reads = "10%"
)
```

Technical note: `min_coverage` and `min_junction_reads` are applied to raw
counts before any library-size normalisation is applied.

## Stranded RNA-seq data

The `strandedness` argument describes how read strand should be interpreted.
It can be one of `"unstranded"`, `"forward"`, or `"reverse"`, and can be
supplied either as a single value for all BAM files or one value per BAM file.
The example BAM files are unstranded, so both read orientations contain signal
from the transcript.

```{r strandedness, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    strandedness = "unstranded"
)
```

Specifying a stranded protocol for unstranded BAM files filters reads by strand.
As a result, either `"reverse"` or `"forward"` shows only part of the coverage
that is visible when `strandedness = "unstranded"`.

```{r forward_strandedness, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    strandedness = "forward"
)
```

```{r reverse_strandedness, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    strandedness = "reverse"
)
```

For paired-end BAM files, `spliceTerrain()` detects paired-end status from the
BAM alignment flags and imports alignments accordingly.
If `region` specifies a strand and `strandedness` is not `"unstranded"`,
alignments are restricted to the resolved region strand.

## Normalising between samples

When samples have different sequencing depths, coverage and junction counts can
be normalised with `lib_size`.
Additional normalisation factors, such as `edgeR` TMM factors, can be supplied
with `norm_factors`.
The effective library size is calculated as `lib_size * norm_factors`.

By default, counts are normalised to the median effective library size.
Use `normalise_to` to set the target explicitly.
The `common_y` argument is useful when comparing samples because it puts sample
panels on the same y-axis scale.

```{r normalisation, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    lib_size = c(40e6, 55e6),
    norm_factors = c(0.95, 1.08),
    normalise_to = 50e6,
    common_y = TRUE
)
```

## Compressed genomic layouts

By default, `compress_introns = TRUE` compacts large uninformative gaps so the
plot focuses on observed signal.
The `intron_width` argument controls the approximate width of compacted introns
in plot space.

```{r compress_introns, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    compress_introns = TRUE,
    intron_width = 500
)
```

If exact genomic distances are more important than compact display, intron
compression can be disabled.

```{r no_compression, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    compress_introns = FALSE
)
```

## Junction arc display

Junction arcs can be tuned with `arc_height`, `arc_side`, and `arc_scale`.
Increasing `arc_height` gives stacked arcs more vertical separation.
The `arc_side` argument controls whether arcs are drawn above, below, or on both
sides of the coverage track.
When `arc_scale = TRUE`, arc line width is scaled by junction read count after
filtering.
Junction label size is controlled with `junc_text_size`.

```{r arcs, fig.wide=TRUE, fig.height=6}
spliceTerrain(
    bam = bams,
    region = region,
    psi = "chr14:70234854-70234854",
    arc_height = 2,
    arc_side = "above",
    arc_scale = TRUE,
    junc_text_size = 2.5
)
```

## Sample colours and panel layout

The `colours` argument controls the colours used for sample coverage and
junction arcs.
It can be a single colour used for all samples or one colour per BAM file.

Axis text sizes can be adjusted with `axis_title_size` and `axis_text_size`.

```{r layout, fig.wide=TRUE, fig.height=7.5}
spliceTerrain(
    bam = bams,
    region = region,
    annotation = annotation,
    colours = c("darkblue", "darkred"),
    panel_heights = c(2, 2, 1),
    axis_title_size = 20,
    axis_text_size = 12
)
```

## Returning and reusing plot data

Most users can work directly with the final plot returned by `spliceTerrain()`.
For advanced workflows, set `return_ctx = TRUE` to return the processed context
instead.
This contains validated settings, imported alignments, summarised coverage,
junction counts, annotation, and overlays.
All coordinate-bearing fields use genomic coordinates; compressed plot
coordinates are generated internally only when the context is plotted.

```{r return_ctx}
ctx <- spliceTerrain(
    bam = bams,
    region = region,
    annotation = annotation,
    min_junction_reads = 1,
    panel_heights = c(2, 2, 1),
    min_arrow = 100,
    return_ctx = TRUE
)

names(ctx)
```

The context can be passed back with the `ctx` argument to redraw the plot
without re-importing BAM files or recalculating coverage and junctions.
For example, the annotation can be used to retain only coverage overlapping
annotated exons and junctions that exactly match annotated introns.

```{r ctx_plot, fig.wide=TRUE, fig.height=6}
ctx$cov <- ctx$cov[overlapsAny(ctx$cov, ctx$annotation)]
ctx$juncs <- ctx$juncs[ctx$juncs$annotation_match]
spliceTerrain(ctx = ctx)
```

The `annotation_match` field records whether each junction exactly matches an
intron implied by the supplied annotation.
Plotting settings can also be changed directly, for example with
`ctx$colours <- c("darkblue", "darkred")`.
Settings that control BAM import, filtering, or normalisation are not reapplied
when a context is replotted because its coverage and junctions are already
processed.

# Session information

```{r session_info}
sessionInfo() |>
    pander()
```
