---
title: "Creating Sashimi-Style Plots with spliceTerrain"
author:
- name: Lachlan Baer
  affiliation:
  - Robinson Research Institute, Adelaide University, Adelaide, SA, Australia
  - College of Health, Adelaide University, Adelaide, SA, Australia
  email: baerlachlan@gmail.com
package: spliceTerrain
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    css: custom.css
# bibliography: references.bib
vignette: >
    %\VignetteIndexEntry{Creating Sashimi-Style Plots with spliceTerrain}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r global-opts, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", message = FALSE, crop = NULL
)
if (!interactive())
    # Wide chunk output will horizontal scroll instead of wrap
    options(width = 999)
```

# Setup

`BiocManager` is recommended for the installation of packages required in this
vignette.
`BiocManager` handles the installation of packages from Bioconductor, CRAN and
GitHub repositories.

```{r install, eval=FALSE}
if (!"BiocManager" %in% rownames(installed.packages()))
    install.packages("BiocManager")
pkgs <- c(
    "spliceTerrain", "RNAseqData.HNRNPC.bam.chr14", "EnsDb.Hsapiens.v75",
    "ensembldb", "stringr", "tibble", "pander"
)
BiocManager::install(pkgs, update = FALSE)
```

Now we can load the required packages.

```{r pkgs}
library(spliceTerrain)
library(RNAseqData.HNRNPC.bam.chr14)
library(EnsDb.Hsapiens.v75)
library(ensembldb)
library(stringr)
library(tibble)
library(pander)
```

# Quick-start

At a minimum, we require BAM file paths and a genomic interval to plot.

Here we use BAM files from the `RNAseqData.HNRNPC.bam.chr14` package as example data, subsetting two of these files to showcase `spliceTerrain`.
In the most simple case, a genomic interval can be specified as a character vector of length 1.
That's all we need to build a simple sashimi plot!

```{r quickstart-2, fig.wide=TRUE, fig.height=6}
bams <- RNAseqData.HNRNPC.bam.chr14_BAMFILES[c(7,1)]
region <- "chr14:70,233,810-70,238,690"
spliceTerrain(bam = bams, region = region)
```

# Introduction

Sashimi plots are a visualisation tool for RNA-seq data that display summarised read coverage across a genomic interval and how splice junctions bridge exons.
By displaying both per-base coverage and junction-spanning reads, sashimi plots provide an intuitive view of transcript structure and enable assessment of exon usage, alternative splicing events, and isoform differences between samples or conditions.

## Motivation

`spliceTerrain` was designed to integrate seamlessly with the Bioconductor ecosystem, utilising core Bioconductor classes and methods at the centre of its workflow.
During RNA-seq analysis, genomic intervals, annotations, and features of interest are often represented as `GRanges` objects.
`spliceTerrain` leverages this to enable straightforward visualisation that integrates seamlessly into `Rmarkdown`/`Quarto` documents, while promoting reproducibility through a user-friendly single-function R interface for generating sashimi-style plots within script-based analyses.
`spliceTerrain` enables publication-quality figures while retaining flexibility.
Users can optionally access the underlying plot data and individual plot panels via the `return_data` argument, allowing further customisation or arrangement of panels.

## Previous works

Several established and wonderful tools also exist for plotting sashimi-style visualisations, such as `MISO`, `ggsashimi` and `IGV`.
`spliceTerrain` complements these by offering a Bioconductor-native, highly configurable approach tailored to bioinformaticians that adopt an `R`-based workflow.

# Complete guide

## Loading required data



## Annotations

```{r}
ensDb <- EnsDb.Hsapiens.v75
```

```{r}
region <- transcripts(ensDb, filter = TxIdFilter("ENST00000394366"))
seqlevelsStyle(region) <- "ucsc"
```

```{r}
annotation <- exonsBy(ensDb, by = "tx", filter = GeneNameFilter("SRSF5"))
seqlevelsStyle(annotation) <- "ucsc"
```

```{r, fig.wide=TRUE, fig.height=9}
spliceTerrain(bam = bams, region = region, annotation = annotation)
```

```{r, fig.wide=TRUE, fig.height=9}
spliceTerrain(
    bam = bams, region = region, annotation = annotation, panel_heights = c(
        1,1,2
    )
)
```

## Panel heights

## Annotation labels

## Colours

## Compressing introns

intron_width

## Calculating PSI

## Highlighting regions

highlight_colour

## Strandedness

## Minimum coverage

## Customizing arcs

arc_height
scale_arcs

## Returning plot data

This way you can arrange the plots as you like for publications, investigate junction level data to find the LSV that you want to calculate PSI values for, troubleshoot, etc.

# Session Information

```{r session_info}
sessionInfo() |>
    pander()
```
